Volatility Metric to Detect Anomalies in Source Code Repositories

### 2020-05-17 paper/article.tex
more text

\documentclass[sigconf]{acmart}
\title{Volatility Metric to Detect Anomalies in Source Code Repositories}
\author{Yegor Bugayenko}{}{}
\email{yegor.bugayenko@huawei.com}
\affiliation{%
  \institution{Huawei Technologies Co., Ltd.}
  \city{Moscow, Russia}
}
\ccsdesc[100]{Software and its engineering~Metrics}
\keywords{Metrics, Software Size, Software Maintainability}
\setcopyright{acmcopyright}
\copyrightyear{2020}
\acmYear{2020}

@@

Git logs provide information about every single change made by every software
developer during the entire course of the project. Using this information
it's possible to measure which files are being modified frequently. On the
other hand, it's also possible to spot files that are rarely modified and may
be considered as abandoned code, which may be considered as a threat
to the maintainability of the entire project.

We introduce a metric to measure the relationship
between the amount of actively modified files and files which stay
in the repository for a long time without modifications.

In order to demonstrate how the metric works we applied it to
\thetotalrepos{} public Java repositories from GitHub and analyzed
its impact on a two other characteristics of each repository:
the number of files in a repository and the size of it in bytes.
Larger repositories shown larger values of the metric.

@@

To our knowledge, the metrics introduced in this research in order
to detect anomalies in source code repository maintenance, has never been
suggested.

@@

Then, the entire interval between $\check{T}$ (the maximum value)
and $\hat{T}$ (the minimum value) is divided to $Z$ (the input parameter of the method)
equivalent groups:

\begin{align}
\delta &= ( \check{T} - \hat{T} ) / Z \\
G &= [g_1, g_2, \dots, g_{Z}] \\
g_j &= \sum_{i=1}^N [ j(\delta-1) < t_i < j\delta ]
\end{align}

@@

The variance $\text{Var}(g)$ is the volatility of the source code. The smaller
the volatility the more cohesive is the repository and the smaller
the amount of the abandonded code inside it.

@@

The volatility metric was calculated for each repository, using the
formula explained above (the value of $Z$ was set to 64).
Then, a few other metrics were collected
for each repository and their values were compared with the volatility.

The Figure~\ref{fig:1} demonstrates the relationship between
the number of files in the repository~($M_1$) and its volatility~($V$). Both
axixes of the graph have logarithmic scales, for the sake of visual
understandability: the difference between the minimum and the maximum values
of the volatility is logarithmically large. It is visually obvious that
repositories with larger number of files tend to have higher values
of the volatility metric.

\begin{figure}[h]
  \input{files-64.tex}
  \caption{The relationship between the number of files in the repository and its volatility ($Z$ is set to 64)}
  \label{fig:1}
\end{figure}

The Figure~\ref{fig:2} demonstrates the relationship between
the logarithm of the size of the Git repository in bytes~($M_2$) and
the logarithm of its volatility~($V$).
It is visually obvious that
binary-size larger repositories tend to have higher values
of the volatility metric.

\begin{figure}[h]
  \input{bytes-64.tex}
  \caption{The relationship between the size of the Git repository in bytes and its volatility ($Z$ is set to 64)}
  \label{fig:2}
\end{figure}

The Figure~\ref{fig:1} demonstrates the same relationship as the Figure~\ref{fig:1},
but the parameter of the volatility formula is set to 32 instead of 64. The
trend of the graph didn't change much.

\begin{figure}[h]
  \input{files-32.tex}
  \caption{The relationship between the number of files in the repository and its volatility ($Z$ is set to 32)}
  \label{fig:3}
\end{figure}

The Figure~\ref{fig:1} demonstrates the same relationship as the Figure~\ref{fig:1},
but this time the parameter of the volatility formula is increased to 128
instead of the original value of 64. The trend of the graph didn't change much.

\begin{figure}[h]
  \input{files-128.tex}
  \caption{The relationship between the number of files in the repository and its volatility ($Z$ is set to 128)}
  \label{fig:4}
\end{figure}

@@

Also, maintenance anomalies is a sign of differences in customer demand
for the source code components. The module with smaller frequency of changes
must be released less frequently than the module that is being edited every day.
Not paying attention to the growing volatility of the repository means
ignoring anomalies and, because of that, redundant releasing of the code
that doesn't require new releases. Moreover, releasing larger software modules
usually means longer build cycles, which are the key failure factor of
continuous delivery, as explained by~\cite{humble2010}. Getting rid of
maintenance anomalies may help reduce build cycles.

It would be beneficial in future researches to analyze the relationship between
volatility and other parameters of source code repositories,
such as
the programming language being used,
the popularity in GitHub (the number of stars and forks),
the total number of lines of code,
the average number of lines of code per file,
the number of Git commit authors,
the total age of the repository,
the number of issues and pull requests in GitHub,
and many others.
Would be interesting to empiricially demonstrate which of these parameters
(or combinations of them)
impact the behavior of volatility and in which direction.

@@

A new source code volatility metric was introduced and applied
to \thetotalrepos{} open source Java repositories from GitHub. It was
empirically demonstrated that larger repositories have higher values
of the volatility metric.

>>>>

\documentclass[anonymous,sigconf,10pt,nonacm=true]{acmart}
\title{Volatility Metric to Detect Anomalies in Source Code Repositories}
\author{Yegor Bugayenko}{}{}
\email{yegor.bugayenko@huawei.com}
\affiliation{%
  \institution{Huawei Technologies Co., Ltd.}
  \city{Moscow, Russia}
}
\ccsdesc[100]{Software and its engineering~Metrics}
\keywords{Metrics, Software Size, Software Maintainability}
% \setcopyright{acmcopyright}
% \copyrightyear{2020}
% \acmYear{2020}

@@

Git logs provide information about every single change made by every software
developer during the entire course of the project. Using this information
it's possible to measure which files are being modified more frequently than others.
On the other hand, it's also possible to spot files that are rarely modified and may
be considered either as \emph{abandoned} or \emph{stable} code.
\citet{fontana2014} claims that the gap between actively modified code
and the code that is stable should be as narrow as possible in a properly
maintained repository.

Indeed, abandoned code may be considered as a threat
to the maintainability of the entire project, because its lifecycle
is not aligned with the lifecycle of active code, which is modified more intensively.
Later, when a modification is required to the abandonded code, a programmer
may have difficulties finding out the principles of design of the code and
making sure new modifications don't break it.

We introduce a Volatility metric to measure the relationship
between the amount of actively modified files and files which stay
in the repository for a long time without modifications.
The hipothesis is that the Volatility is related to the size of the
repository: the larger the repository the higher the volatility. Larger
repositories are an obvious threat to code maintainability. Thus, if
the relationship between size and Volatility exists, the new metric
can be used to identify code anomalies and detect badly maintained repositories.

In order to validate the hipothesis and demonstrate how the metric works we applied it to
\thetotalrepos{} public Java repositories from GitHub and analyzed
its impact on two other characteristics of each repository:
the number of files in a repository and the size of it in bytes.

Larger repositories shown larger values of the metric.

@@

\citet{fontana2014} suggested to measure ``File Volatility'' as the ratio
between the maximum number of changes happened to a file line and the number
of all changes to the file since its creation. Going further, they also suggested to
calculate ``Repository Stability'' metric as a relationship between
active files (with high File Volatility) and closed files.
The also concluded that repositories where most files change
over time can suffer from organizational or design issues.

To our knowledge, the metric introduced in this research, in order
to detect anomalies in source code repository maintenance, has never been
suggested.

@@

Here, $t_i$ is the number of commits found in the repository, which
modified the $i$-th file. Then, the entire interval between $\check{T}$ (the maximum value)
and $\hat{T}$ (the minimum value) is divided to $Z$ (the input parameter of the method)
equivalent groups:

\begin{align}
\delta &= ( \check{T} - \hat{T} ) / Z \\
G &= [g_1, g_2, \dots, g_{Z}] \\
g_j &= \sum_{i=1}^N [ j \times (\delta-1) < t_i < j \times \delta ]
\end{align}

Here, $g_j$ is the total number of $t_i$, which are larger than
$j \times (\delta-1)$ and smaller than $j \times \delta$. In other words,
$T$, the entire set of measurements, is sorted and then split into $Z$ sectors,
where $g_j$ is the counter of measurements that belong to the sector $j$.

@@

The variance $\text{Var}(g)$ is the Volatility of the source code. The smaller
the Volatility the more cohesive is the repository and the smaller
the amount of the abandonded code inside it. The metric demonstrates how
large is the gap between intensively modified files and the abandoned ones:
the larger the variance, the wider the graph, and the larger the gap.
A repository with a thousand files all being modified at the same pace would
have a Volatility of zero. On the other hand, a repository with just a few
files, where one of them is modified a thousand times, while others are modified
just once, will have the Volatility of a miillion.

@@

The Volatility metric was calculated for each repository, using the
formula explained above (the value of $Z$ was set to 64).
Then, a few other metrics were collected
for each repository and their values were compared with the Volatility.

The Figure~\ref{fig:1} demonstrates the relationship between
the number of files in the repository~($M_1$) and its Volatility~($V$).
Both axixes of the graph have logarithmic scales, for the sake of visual
understandability: the difference between the minimum and the maximum values
of the Volatility is logarithmically large. It is visually obvious that
repositories with larger number of files tend to have higher values
of the Volatility metric.

\begin{figure}[h]
  \input{files-64.tex}
  \caption{The relationship between the number of files in the repository and its Volatility ($Z$ is set to 64)}
  \label{fig:1}
\end{figure}

The Figure~\ref{fig:2} demonstrates the relationship between
the logarithm of the size of the Git repository in bytes~($M_2$) and
the logarithm of its Volatility~($V$).
It is visually obvious that
binary-size larger repositories tend to have higher values
of the Volatility metric.

\begin{figure}[h]
  \input{bytes-64.tex}
  \caption{The relationship between the size of the Git repository in bytes and its Volatility ($Z$ is set to 64)}
  \label{fig:2}
\end{figure}

The Figure~\ref{fig:1} demonstrates the same relationship as the Figure~\ref{fig:1},
but the parameter of the Volatility formula is set to 32 instead of 64. The
trend of the graph didn't change much.

\begin{figure}[h]
  \input{files-32.tex}
  \caption{The relationship between the number of files in the repository and its Volatility ($Z$ is set to 32)}
  \label{fig:3}
\end{figure}

The Figure~\ref{fig:1} demonstrates the same relationship as the Figure~\ref{fig:1},
but this time the parameter of the Volatility formula is increased to 128
instead of the original value of 64. The trend of the graph didn't change much.

\begin{figure}[h]
  \input{files-128.tex}
  \caption{The relationship between the number of files in the repository and its Volatility ($Z$ is set to 128)}
  \label{fig:4}
\end{figure}

@@

Also, maintenance anomalies is a sign of differences in customer demand
for the source code components. The module with smaller frequency of changes
must be released less frequently than the module that is being edited every day.
Not paying attention to the growing Volatility of the repository means
ignoring anomalies and, because of that, redundant releasing of the code
that doesn't require new releases. Moreover, releasing larger software modules
usually means longer build cycles, which are the key failure factor of
continuous delivery, as explained by~\cite{humble2010}. Getting rid of
maintenance anomalies may help reduce build cycles.

It would be beneficial in future researches to analyze the relationship between
Volatility and other parameters of source code repositories,
such as
the programming language being used,
the popularity in GitHub (the number of stars and forks),
the total number of lines of code,
the average number of lines of code per file,
the number of Git commit authors,
the total age of the repository,
the number of issues and pull requests in GitHub,
and many others.
Would be interesting to empiricially demonstrate which of these parameters
(or combinations of them)
impact the behavior of Volatility and in which direction.

@@

A new source code Volatility metric was introduced and applied
to \thetotalrepos{} open source Java repositories from GitHub. It was
empirically demonstrated that larger repositories have higher values
of the Volatility metric.
